---
title: "Assignment 2"
format: pdf
editor: visual
---

```{r, warning=FALSE, message=FALSE}
#| echo: false
dat = read.table("Collider_Data_2025.txt", h = TRUE,stringsAsFactors =TRUE)
dim(dat)
head(dat)
#doesn't want the libraries in code
library(ggplot2)
library(dplyr)
library(r2symbols)
```

# a)

```{r, warning=FALSE, message=FALSE}
#| echo: true
dat$response <- apply(dat[, c("Y1", "Y2", "Y3")], 1, function (x){
  if (x[1] == 1) return("code-Alpha")
  if (x[2] == 1) return("code-Beta")
  if (x[3] == 1) return("code-Rho")
})
ggplot(dat, aes(x=X1, y=X2, color=response ))+geom_point(size=2)+ggtitle("Scatter plot of Collider Data Feature Space")+ theme_minimal()



```

The scatter plot shows how the particles are distributed in the X1-X2 plane. We can see that the process creates a complex and non-linear separation in the points plotted. Therefore, the use of a neural network is justified as they are known for their ability to learn non-linear decision boundaries which is presented in the scatter plot.

# b)

```{r, warning=FALSE, message=FALSE}
#| echo: true
softmax <- function(Z) 
  {
 
  Z_shift <- Z - matrix(apply(Z, 2, max),          # subtract column-wise maxima
                        nrow = 3, ncol = ncol(Z), 
                        byrow = TRUE)
  
  expZ    <- exp(Z_shift)                          # exponentiate
  denom   <- matrix(colSums(expZ),                 # column-wise sums
                    nrow = 3, ncol = ncol(Z), 
                    byrow = TRUE)
  
  expZ / denom                                     # element-wise division
  }


```

# c)

$$
\begin{aligned}
\texttt{calc_Ci <- function(y_true, y_pred) \{}
&\quad \texttt{if (y_true == 1)\{} \\
&\qquad \texttt{return(-log(y_pred))} \\
&\quad \texttt{\}} \\
&\quad \texttt{else \{} \\
&\qquad \texttt{return(0)} \\
&\quad \texttt{\}} \\
\texttt{\}}
\end{aligned}
$$

Evaluating only the component corresponding to the actual class simplifies calculations and enhances numerical stability by avoiding evaluating terms that are known to be 0.

# d)

```{r, warning=FALSE, message=FALSE}
#| echo: true
g <- function(Yhat, Y, eps = 1e-15) {
  # Yhat, Y : N × q matrices   (rows = obs, cols = classes)
  N <- nrow(Y)
  -sum( Y * log( pmax(Yhat, eps) ) ) / N
}

```

# e)

number of parameters = 2p^2^+2p+2pm+2m+m^2^+mq+q

# f)

```{r, warning=FALSE, message=FALSE}
#| echo: true

af_forward <- function(X, Y, theta, m, nu)
{
  N <- nrow(X)
  p <- ncol(X)
  q <- ncol(Y)
  
  index <- 1:(2(p^2)) #W1 : p(p+p)
  W1 <- matrix(theta[index], nrow=p)
  
  index <- max(index)+1:(2*p) #b1 : (p+p)
  b1 <- theta[index]
  
  index <- max(index)+1:((2*p)*m) #W2 : (p+p)*m
  W2 <- matrix(theta[index], nrow=2*p)
  
  index <- max(index)+1:m #b2 : m
  b2 <- theta[index]
  
  index <- max(index)+1:(m*m) #W3 : (m*m)
  W3 <- matrix(theta[index], nrow=m)
  
  index <- max(index)+1:m #b3 : m
  b3 <- theta[index]
  
  index <- max(index)+1:(m*q) #W4 : (m*q)
  W4 <- matrix(theta[index], nrow=m)
  
  index <- max(index)+1:q #b4 : q
  b4 <- theta[index]
  
  #softmax function
  softmax <- function(Z) 
    {
  
  Z_shift <- Z - matrix(apply(Z, 2, max),          # subtract column-wise maxima
                        nrow = 3, ncol = ncol(Z), 
                        byrow = TRUE)
  
  expZ    <- exp(Z_shift)                          # exponentiate
  denom   <- matrix(colSums(expZ),                 # column-wise sums
                    nrow = 3, ncol = ncol(Z), 
                    byrow = TRUE)
  
  expZ / denom                                     # element-wise division
  }
  
  #forward propagation 
  H1 <- tanh( X  %*% W1 + matrix(b1, N, 2*p, TRUE) )       # aug-layer                           
  H2 <- tanh( H1 %*% W2 + matrix(b2, N, m, TRUE) )      # 2nd hidden
  H3 <- tanh(H2 %*% W3 + matrix(b3, N, m, TRUE))
  Z <- H3 %*% W4 + matrix(b4, N, q, TRUE)          # logits
  probs <- softmax(Z)                                          # softmax

  #losses & objective  
  loss <- g(probs, Y)                 # cross-entropy
  obj  <- loss + (nu / 2) * sum(theta^2)
  list(probs = probs, loss = loss, obj = obj)

}

```
